### 2023-02-13 chapters/fairness/justicia/introduction.tex
addressing reviewers comments
highlevel pass

\begin{figure}[!t]
	\centering
	\subfloat[Dependency among features and prediction]{\scalebox{1}{	
			\begin{tikzpicture}[x=1cm,y=0.6cm]
			% Define nodes
			\node[latent,scale=1] (a1) {$\textrm{age}$} ; %
			\node[obs, scale=1, below=of a1, xshift=-.8cm] (h) {$\textrm{fitness}$}; %
			\node[obs, scale=1, below=of a1, xshift=.8cm] (i) {$\textrm{income}$}; %
			\node[obs, scale=1, below=of h, xshift=.8cm] (p) {$\widehat{Y}$}; %			
			%%add edge
			\edge[] {a1} {h,i} ;
			\edge[] {h,i} {p} ;
			\end{tikzpicture}
	}\label{fairness_justicia_fig:fair_example_dependency}}\hfil
	\subfloat[Age-dependent distributions of non-sensitive features]{
		\includegraphics[scale=0.45]{figures/fairness/fif/sanity_distribution}
	\label{fairness_justicia_fig:fair_example_distribution}}

	\subfloat[Trained decision tree]{
		\scalebox{0.55}{	
			\begin{tikzpicture}[x=1cm,y=1.8cm]
			\node [box, scale=1.5]                                    (p)      {fitness $\geq 0.61$};
			\node [scale=1.5, box, below= of p, xshift=-2.1cm, yshift=1.2cm]    (a1)    {income\\ $\geq 0.29$};
			\node [scale=1.5, box, below= of p, xshift=2.1cm, yshift=1.2cm]     (a2)    {income $\geq 0.69$};
			\node [scale=1.5,below= of a1, xshift=-1.5cm, yshift=0.8cm]  (a11)    { $\widehat{Y}= 1$};
			\node [scale=1.5,below= of a1, xshift=1.5cm, yshift=0.8cm]   (a12)    { $\widehat{Y}=0 $};
			\node [scale=1.5,below= of a2, xshift=-1.5cm, yshift=0.8cm]  (a21)    { $\widehat{Y}= 1$};
			\node [scale=1.5,below= of a2, xshift=1.5cm, yshift=0.8cm]  (a22)    { $\widehat{Y}= 0$};
			%
			\path [line] (p) -|         (a1) node [scale=1.5,midway, above]  {Y};
			\path [line] (p) -|         (a2) node [scale=1.5,midway, above]  {N};
			\path [line] (a1) -|       (a11) node [scale=1.5,midway, above]  {Y};
			\path [line] (a1) -|       (a12) node [scale=1.5,midway, above]  {N};
			\path [line] (a2) -|       (a21) node [scale=1.5,midway, above]  {Y};
			\path [line] (a2) -|       (a22) node [scale=1.5,midway, above]  {N};
			\end{tikzpicture}}
	\label{fairness_justicia_fig:fair_example_dt}}\hfil
	\subfloat[Group specific conditional probability of the classifier's positive prediction]{
		\scalebox{1}{
			\begin{tikzpicture}
			\begin{axis}[
			enlarge x limits=0.2,
			symbolic x coords={$ \text{age} \ge 40 $, $ \text{age} < 40 $},
			xtick={$ \text{age} \ge 40 $, $ \text{age} < 40 $},  
%			xticklabel style={text height=2ex}, 
			ylabel={$ \Pr[\hat{Y} =  1] $},
			xlabel={Sensitive groups},
			ymin=0,
			width= 0.3\textwidth,]

@@

\subsubsection{Fair Machine Learning} Statistical discrimination caused by classifiers has motivated researchers to formulate several definitions of fairness and develop associated algorithms to mitigate bias. In this chapter, we focus on a popular concept of fairness, known as group fairness. Existing group fairness metrics mostly belong to three categories: \textit{independence}, \textit{separation}, and \textit{sufficiency}~\cite{mehrabi2019survey}. Independence metrics, such as demographic parity, statistical parity, and group parity, try and ensure the outcomes of a classifier to be independent of the groups that the individuals belong to~\cite{feldman2015certifying,dwork2012fairness}. Separation metrics, such as equalized odds, define a classifer to be fair if the probability of getting the same outcomes for different groups are same, separately for data from each possible ground-truth class label~\cite{hardt2016equality}. Sufficiency metrics, such as counterfactual fairness, constrain the probability of outcomes to be independent of individual's sensitive data given their identical non-sensitive data~\cite{kusner2017counterfactual}.

In Figure~\ref{fairness_justicia_fig:fair_example}, independence is satisfied if the probability of getting insurance is same for both the age groups. Separation is satisfied if the number of `actually' (ground-truth) ineligible and eligible people getting the insurance are same. Sufficiency is satisfied if the eligibility is independent of their age given their features are the same. Thus, we see that the metrics of fairness can be contradictory and complimentary depending on the application and the data~\cite{corbett2018measure}. To this end, different algorithms have also been devised to ensure one or multiple of the fairness definitions. These algorithms try to rectify and mitigate the bias in the data and thus in the classifier in three ways: \textit{pre-processing} the data~\cite{kamiran2012data,zemel2013learning,calmon2017optimized}, \textit{in-processing} the classifier~\cite{zhang2018mitigating}, and \textit{post-processing} the outcomes of a classifier~\cite{kamiran2012decision,hardt2016equality}.

\subsubsection{Fairness Verifiers} Due to the abundance of fairness metrics and difference in algorithms to achieve them, it has become necessary to verify different fairness metrics over datasets and algorithms.

@@

\subsubsection{Our Contribution.} From this vantage point, \textit{we propose to model verifying different fairness metrics as a Stochastic Boolean Satisfiability (SSAT) problem}~\cite{littman2001stochastic}. SSAT was originally introduced by ~\cite{papadimitriou1985games} to model {\em games against nature}. In this work, we primarily focus on reductions to the exist-random quantified fragment of SSAT, which is also known as E-MAJSAT~\cite{littman2001stochastic}.   SSAT is a conceptual framework that has been employed to capture several fundamental problems in artificial intelligence such as the computation of maximum a posteriori (MAP) hypothesis~\cite{fremont2017maximum},  propositional probabilistic planning~\cite{majercik2007appssat},  and circuit verification~\cite{lee2018towards}. Furthermore, our choice of SSAT as a target formulation is motivated by the recent algorithmic progress that has yielded efficient SSAT tools~\cite{lee2017solving,lee2018solving}.

>>>>

\begin{figure}[!t]
	\centering
	\subfloat[Dependency among features and prediction]{\scalebox{1}{	
			\begin{tikzpicture}[x=1cm,y=0.6cm]
			% Define nodes
			\node[latent,scale=1] (a1) {$\textrm{age}$} ; %
			\node[obs, scale=1, below=of a1, xshift=-.8cm] (h) {$\textrm{fitness}$}; %
			\node[obs, scale=1, below=of a1, xshift=.8cm] (i) {$\textrm{income}$}; %
			\node[obs, scale=1, below=of h, xshift=.8cm] (p) {$\widehat{Y}$}; %			
			%%add edge
			\edge[] {a1} {h,i} ;
			\edge[] {h,i} {p} ;
			\end{tikzpicture}
	}\label{fairness_justicia_fig:fair_example_dependency}}\hfill
	\subfloat[Age-dependent distributions of non-sensitive features]{
		\includegraphics[scale=0.45]{figures/fairness/fif/sanity_distribution}
	\label{fairness_justicia_fig:fair_example_distribution}}

	\subfloat[Trained decision tree]{
		\scalebox{0.55}{	
			\begin{tikzpicture}[x=1cm,y=1.8cm]
			\node [box, scale=1.5]                                    (p)      {fitness $\geq 0.61$};
			\node [scale=1.5, box, below= of p, xshift=-2.1cm, yshift=1.2cm]    (a1)    {income\\ $\geq 0.29$};
			\node [scale=1.5, box, below= of p, xshift=2.1cm, yshift=1.2cm]     (a2)    {income $\geq 0.69$};
			\node [scale=1.5,below= of a1, xshift=-1.5cm, yshift=0.8cm]  (a11)    { $\widehat{Y}= 1$};
			\node [scale=1.5,below= of a1, xshift=1.5cm, yshift=0.8cm]   (a12)    { $\widehat{Y}=0 $};
			\node [scale=1.5,below= of a2, xshift=-1.5cm, yshift=0.8cm]  (a21)    { $\widehat{Y}= 1$};
			\node [scale=1.5,below= of a2, xshift=1.5cm, yshift=0.8cm]  (a22)    { $\widehat{Y}= 0$};
			%
			\path [line] (p) -|         (a1) node [scale=1.5,midway, above]  {Y};
			\path [line] (p) -|         (a2) node [scale=1.5,midway, above]  {N};
			\path [line] (a1) -|       (a11) node [scale=1.5,midway, above]  {Y};
			\path [line] (a1) -|       (a12) node [scale=1.5,midway, above]  {N};
			\path [line] (a2) -|       (a21) node [scale=1.5,midway, above]  {Y};
			\path [line] (a2) -|       (a22) node [scale=1.5,midway, above]  {N};
			\end{tikzpicture}}
	\label{fairness_justicia_fig:fair_example_dt}}\hfill
	\subfloat[Group specific conditional probability of the classifier's positive prediction]{
		\scalebox{1}{
			\begin{tikzpicture}
			\begin{axis}[
			enlarge x limits=0.2,
			symbolic x coords={$ \text{age} \ge 40 $, $ \text{age} < 40 $},
			xtick={$ \text{age} \ge 40 $, $ \text{age} < 40 $},  
%			xticklabel style={text height=2ex}, 
			ylabel={$ \Pr[\hat{Y} =  1] $},
			xlabel={Sensitive groups},
			ymin=0,
			width= 0.3\textwidth,]

@@

\paragraph{Fairness in Machine Learning.} Statistical discrimination caused by classifiers has motivated researchers to formulate several definitions of fairness and develop associated algorithms to mitigate bias. In this chapter, we focus on a popular concept of fairness, known as group fairness. Existing group fairness metrics mostly belong to three categories: \textit{independence}, \textit{separation}, and \textit{sufficiency}~\cite{mehrabi2019survey}. Independence metrics, such as demographic parity, statistical parity, and group parity, try and ensure the outcomes of a classifier to be independent of the groups that the individuals belong to~\cite{feldman2015certifying,dwork2012fairness}. Separation metrics, such as equalized odds, define a classifer to be fair if the probability of getting the same outcomes for different groups are same, separately for data from each possible ground-truth class label~\cite{hardt2016equality}. Sufficiency metrics, such as counterfactual fairness, constrain the probability of outcomes to be independent of individual's sensitive data given their identical non-sensitive data~\cite{kusner2017counterfactual}.

In Figure~\ref{fairness_justicia_fig:fair_example}, independence is satisfied if the probability of getting insurance is same for both the age groups. Separation is satisfied if the number of `actually' (ground-truth) ineligible and eligible people getting the insurance are same. Sufficiency is satisfied if the eligibility is independent of their age given their features are the same. Thus, we see that the metrics of fairness can be contradictory and complementary depending on the application and the data~\cite{corbett2018measure}. To this end, different algorithms have also been devised to ensure one or multiple of the fairness definitions. These algorithms try to rectify and mitigate the bias in the data and thus in the classifier in three ways: \textit{pre-processing} the data~\cite{kamiran2012data,zemel2013learning,calmon2017optimized}, \textit{in-processing} the classifier~\cite{zhang2018mitigating}, and \textit{post-processing} the outcomes of a classifier~\cite{kamiran2012decision,hardt2016equality}.

\paragraph{Fairness Verifiers.} Due to the abundance of fairness metrics and difference in algorithms to achieve them, it has become necessary to verify different fairness metrics over datasets and algorithms.

@@

\paragraph{Contribution.} From this vantage point, \textit{we propose to model verifying different fairness metrics as a Stochastic Boolean Satisfiability (SSAT) problem}~\cite{littman2001stochastic}. SSAT was originally introduced by ~\cite{papadimitriou1985games} to model {\em games against nature}. In this work, we primarily focus on reductions to the exist-random quantified fragment of SSAT, which is also known as E-MAJSAT~\cite{littman2001stochastic}.   SSAT is a conceptual framework that has been employed to capture several fundamental problems in artificial intelligence such as the computation of maximum a posteriori (MAP) hypothesis~\cite{fremont2017maximum},  propositional probabilistic planning~\cite{majercik2007appssat},  and circuit verification~\cite{lee2018towards}. Furthermore, our choice of SSAT as a target formulation is motivated by the recent algorithmic progress that has yielded efficient SSAT tools~\cite{lee2017solving,lee2018solving}.

### 2023-02-13 chapters/fairness/justicia/practical.tex
with new formatting
highlevel pass

\subsubsection{Extending {\justicia} to Decision Trees and Linear Classifiers. }
In the SSAT approach, we assume that the classifier $\alg$ is represented as a CNF formula.  
We extend {\justicia} beyond CNF classifiers to decision trees and linear classifiers, which are widely used in the fairness studies~\cite{zemel2013learning,raff2018fair,zhang2019faht}.
%{\color{red}In the literature of interpretable machine learning, several studies have been conducted for learning CNF classifiers in the supervised learning setting, which include but are not limited to the work of~\cite{angelino2017learning,malioutov2018mlic,ghosh19incremental}.  We leverage these techniques. (not needed)}

@@

\subsubsection{Extending to Continuous Features.}
In practical problems, features are generally real-valued or categorical but classifiers, which are naturally expressed as CNF such as~\cite{GMM20}, are generally trained on a Boolean abstraction of input features. In order to perform the Boolean abstraction, each categorical feature is one-hot encoded and each real-valued feature is discretized into a set of Boolean features~\cite{LKCL2019,GMM20}.

>>>>

\subsubsection{Extension to Decision Trees and Linear Classifiers}
In the SSAT approach, we assume that the classifier $\alg$ is represented as a CNF formula.  
We extend {\justicia} beyond CNF classifiers to decision trees and linear classifiers, which are widely used in the fairness studies~\cite{zemel2013learning,raff2018fair,zhang2019faht}.
%{\color{red}In the literature of interpretable machine learning, several studies have been conducted for learning CNF classifiers in the supervised learning setting, which include but are not limited to the work of~\cite{angelino2017learning,malioutov2018mlic,ghosh19incremental}.  We leverage these techniques. (not needed)}

@@

\subsubsection{Extension to Continuous Features}
In practical problems, features are generally real-valued or categorical but classifiers, which are naturally expressed as CNF such as~\cite{GMM20}, are generally trained on a Boolean abstraction of input features. In order to perform the Boolean abstraction, each categorical feature is one-hot encoded and each real-valued feature is discretized into a set of Boolean features~\cite{LKCL2019,GMM20}.

### 2023-02-13 chapters/interpretability/imli/incremental_learning.tex
highlevel pass

\section{Incremental Learning with MaxSAT-based formulation}
\label{interpretability_imli_sec:incremental_learning}
The complexity of the MaxSAT query in Section~\ref{interpretability_imli_sec:encoding} increases with the number of samples in the training dataset and the number of clauses to be learned in a CNF formula. In this section, we discuss an incremental learning technique that along with the MaxSAT formulation can generate classification rules efficiently. Our incremental learning is built on two concepts: (i)
mini-batch learning and (ii) iterative learning. We discuss both in detail in the following.

@@

\paragraph{A Heuristic for Mini-Batch Learning.} Let $ \Rule' $ be a classifier learned on the previous batch. In mini-batch learning, we aim to learn a new classifier $ \Rule $ that can generalize to both the current batch and previously seen samples. For that, we consider a soft constraint such that $ \Rule $ does not \emph{differ much} from $ \Rule' $ while training on the current batch. Thus, we hypothesize that by constraining $ \Rule $ to be syntactically similar to $ \Rule' $, it is possible to generalize well; because samples in all batches originate from the same distribution\footnote{We propose Hamming distance heuristics in {\imli} with the assumption that distribution remains same across batches. One way to account for distribution shifts is to consider last $ p $  $ (> 1) $ batches instead of the (single) previous batch in the objective function in mini-batch learning. For a feature variable $ B^i_j $, we consider its majority assignment in last $ p $ classification rules and encode as a soft clause to retain the majority assignment in the current batch. Moreover, we can reweigh the soft clause by prioritizing assignments of $ B^i_j $ in recent batches.}. Since our study focuses on rule-based classifiers, we consider the Hamming distance between two classifiers as a notion of their syntactic dissimilarity. In the following, we define the Hamming distance between two CNF classifiers $ \Rule $, $ \Rule' $ with the same number of clauses as follows.
\[
	d_{H}(\Rule, \Rule') = \sum_{i=1}^k \Big(\sum_{v\in C_i} \mathds{1}(v \not\in C'_i) +  \sum_{v\in C'_i} \mathds{1}(v \not\in C_i) \Big), 
\]

>>>>

\section{Incremental Learning with MaxSAT}
\label{interpretability_imli_sec:incremental_learning}
The complexity of the MaxSAT query in Section~\ref{interpretability_imli_sec:encoding} increases with the number of samples in the training dataset and the number of clauses to be learned in a CNF formula. In this section, we discuss an incremental learning technique that along with the MaxSAT formulation can generate classification rules efficiently. Our incremental learning is built on two concepts: (i)
mini-batch learning and (ii) iterative learning. We discuss both in detail in the following.

@@

\paragraph{A Heuristic for Mini-Batch Learning.} Let $ \Rule' $ be a classifier learned on the previous batch. In mini-batch learning, we aim to learn a new classifier $ \Rule $ that can generalize to both the current batch and previously seen samples. For that, we consider a soft constraint such that $ \Rule $ does not \emph{differ much} from $ \Rule' $ while training on the current batch. Thus, we hypothesize that by constraining $ \Rule $ to be syntactically similar to $ \Rule' $, it is possible to generalize well; because samples in all batches originate from the same distribution\footnote{We propose Hamming distance heuristics in {\imli} with the assumption that the probability distribution of features remains same across batches. One way to account for distribution shifts is to consider last $ p $  $ (> 1) $ batches instead of the (single) previous batch in the objective function in mini-batch learning. For a feature variable $ B^i_j $, we consider its majority assignment in last $ p $ classification rules and encode as a soft clause to retain the majority assignment in the current batch. Moreover, we can reweigh the soft clause by prioritizing assignments of $ B^i_j $ in recent batches.}. Since our study focuses on rule-based classifiers, we consider the Hamming distance between two classifiers as a notion of their syntactic dissimilarity. In the following, we define the Hamming distance between two CNF classifiers $ \Rule $, $ \Rule' $ with the same number of clauses as follows.
\[
	d_{H}(\Rule, \Rule') = \sum_{i=1}^k \Big(\sum_{v\in C_i} \mathds{1}(v \not\in C'_i) +  \sum_{v\in C'_i} \mathds{1}(v \not\in C_i) \Big), 
\]

### 2023-02-13 chapters/interpretability/imli/problem_formulation.tex
notations updated
highlevel pass

 \section{Problem Formulation}
 \label{interpretability_imli_sec:problem}
Given 
\begin{enumerate}
	\item a dataset $ \dataset = \{(\mathbf{x}^{(i)}, y^{(i)})\}_{i=1}^n$ of $ n $ samples, where feature vector $ \mathbf{x}^{(i)} \in \{0, 1\}^m $ contains $ m $ features and class label $ y^{(i)} \in \{0,1\} $,
	\item a positive integer $ k \ge 1 $ denoting the number of clauses to be learned in the classification rule, and
	\item a regularization parameter $ \lambda \in \mathbb{R}^+ $,
\end{enumerate}
we learn a classifier $ \Rule $ represented as a $ k $-clause CNF formula that can separate samples of class $ 1 $ from class $ 0 $.

 Our goal is to learn classifiers that balance two 
 goals: of  being accurate but also interpretable.  
 Various notions of interpretability have been proposed in the context of   classification problems. A common proxy for interpretability in the context of decision rules 
 is the sparsity of rule. For instance, a rule involving fewer literals is highly interpretable.  In this work, we minimize the total number of literals in all clauses, which motivates us to  find $ \Rule  $ with minimum  $ |\Rule| $. Let $ \Rule $ classify all samples correctly during training. Among all the classification rules that classify all samples correctly,  we choose the sparsest (most interpretable) such $ \Rule $.

>>>>

 \section{Problem Formulation}
 \label{interpretability_imli_sec:problem}
We are given 
\begin{enumerate}
	\item a dataset $ \dataset = \{(\mathbf{x}^{(i)}, y^{(i)})\}_{i=1}^n$ of $ n $ samples, where feature vector $ \mathbf{x}^{(i)} \in \{0, 1\}^m $ contains $ m $ features and class label $ y^{(i)} \in \{0,1\} $,
	\item a positive integer $ k \ge 1 $ denoting the number of clauses to be learned in the classification rule, and
	\item a regularization parameter $ \lambda \in \mathbb{R}^+ $.
\end{enumerate}
Our goal is to learn a rule-based classifier $ \Rule $ represented as a $ k $-clause CNF formula separating samples of class $ 1 $ from class $ 0 $.

 We learn classifiers that balance two 
 goals: of  being accurate but also interpretable.  
 Various notions of interpretability have been proposed in the context of   classification problems. A common proxy for interpretability in the context of decision rules 
 is the sparsity of rule. For instance, a rule involving fewer literals is highly interpretable.  In this work, we minimize the total number of literals in all clauses, which motivates us to  find $ \Rule  $ with minimum  $ |\Rule| $. Let $ \Rule $ classify all samples correctly during training. Among all the classification rules that classify all samples correctly,  we choose the sparsest (most interpretable) such $ \Rule $.
